---
title: "User instructions"

date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User instructions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Purpose of this package
The ```natronbatchupload``` R-package has been developed for employees at the University Museum of the Norwegian University of Science and Technology (NTNU), and it serves to ease and standardize the process of uploading datasets to the NaTRON database. NaTRON is the internal database for the NTNU University Museum. It stores natural history data, including biodiversity and ecological datasets. NaTRON is therefore different from the MUSIT system which store mostly data on objects in the museum collections. All natural history data to be preserved should be uploaded to the NaTRON database, and the database have a generalized structure in order to support most datatypes (botanical, zoological, collection objects, non-biological environmental data, GIS-data). 

This package was built to import location-event data, and is in its current version not designed for large datasets containing repeated measurements (e.g data from dataloggers, biotelemetric studies etc.).

## Content
The process of uploading data to NaTRON using natronbatchupload consist of five steps:

* **Step 1. Restructure dataset**
* **Step 2. Connect to NaTRON** 
* **Step 3. Make location table** 
  * **Step 3.1. View the locations on a map**
  * *Step 3.2 (optional pathway). Find existing NaTRON locations*
    * *Step 3.2.2. Plot NaTRON locations agains original locations*
    * *Step 3.2.3. Modify location table*
* **Step 4. Upload locations**
* **Step 5. Structure event data** 
* **Step 6. Upload event data**
* **Step 7. Structure occurence data** 
* **Step 8. Upload occurrence data to NaTron**
* **Examples of correct data structure**

## Step 1. Restructure
The first step to getting your data on to NaTRON is to restructure the dataset from wide to long format to be compatible with the functions in the ```natronbatchupload``` R-package. Most datasets are punched in Excel in a wide format (ie lots of columns and few rows; see examples at the bottom of this vignette), but NaTRON and most other databases prefer the long format. This step is not covered by this package and must be done manually. We recoment looking into ```reshape2``` and ```dplyr``` for this type of data wrangling. Your dataset should look like our example dataset:
```{r, eval=FALSE}
data("setesdal")
head(setesdal)

```







## Step 2. Connect
Next you need to establish a connection to the NaTRON database. Using the `natron_connect` function makes it easy to connect to NaTRON in R. The functions takes two arguments -- your user name and the database (defults to the sandbox). A pop-up window will appear where you are asked for your password. Save your connection to the environment so that it can be used in subsequent functions. Example of use:

```{r, eval=FALSE}
myConnection <- natron_connect(username)

``` 
Talk to Marc Daverdin about NaTRON access. 







## Step 3. Make location table
Make a dataframe with your unique locations (ie unique coordinates) and additional location specific metedata such as elevation, varbatim coordinates and place names. The returned dataset is structure like the NaTRON table 'Locations' so that it can easily be uploaded later.
```{r, eval=FALSE}
myLocationTable <- location_table(data = setesdal, conn = MyConnection)

```

This step includes a manual check of correct mapping, ie that all the columns that you want to have in the location table is mapped/matched to the correct NaTRON (DwC) terms. 



## Step 3.1. View the locations on a map
Plotting the coordinates an a map help us check that they are correctly entered. 
```{r, eval=FALSE}
map_locations(data = myLocationTable)

# Or plot it vertically like this:
map_locations(data = myLocationTable, vertical = T)

```






### Step 3.2 (optional). Find existing NaTRON locations
If you know or suspect that the locations in your data already exists in NaTRON then you probably want to reuse these locations and the unique locationIDs so that you don't break up any time series data for example. 
`radius_scan()` takes as input a dataframe that contains a minimum of two columns: decimalLatitude and decimalLongitude. It then looks for coordinates in the NaTRON location table within the radius you define and returns them in another dataframe where your original coordinates and locations names are to right. The distance to the closest NaTRON location is given in km. 

```{r, eval=FALSE}
radius <- 8000 #setting it very high here just to make a point
scan <- radius_scan(locationTable = myLocationTable, conn = conn, radius = radius)


``` 


#### Step 3.2.2. Plot NaTRON locations agains original locations

If you have done step 3.2 you can also use \code{map_locations()} to visually check the proximity of your locations and the existing NaTRON locations:

```{r, eval=FALSE}
# Use the 'compare' argument:
map_locations(data = myLocationTable, compare = scan)
```



#### Step 3.2.3. Modify location table
*THIS FUNCTIONALITY IS NOT YET AVAILABLE*

If some or all of your locations already exist in NaTRON (typicall for time series data) then you should idealy not upload them again to NaTRON. Thats a waste of storage space and it makes it harder to join together data from the same locations. However, the following functions that create event and occurence tables needs to link to the NaTRON locations somehow. One term (column) that could be used  to link these tables together is *locationID* which consists of UUIDs. These are not human friendly values and so we need an automated but safe way of harnessing them from NaTRON . \code{radius_scan()} is one way perhaps, but it has proven difficult to make it reliable.


## Step 4. Upload locations 
All locations must be present on NaTRON before the 'real data' is uploaded and you will not be able to upload event-data that don't have a corresponding *locationID* in the database. The link between the 'real data' (ie event data or occurence data) and the location specific metadata is ensured by the relational nature of NaTRON. 


The `upsert_location` function will let you upload new locations to the NaTRON database. The format of the data to be uploaded must exactly the same as the NaTRON database table. Example of use:

```{r, eval = F}
upsert_locations(myLocationTable, conn = myConnection)

```



## 5. Structure event-data

This step organises your event data table to be compatible and ready to upload to NaTRON. The event data table needs to look just like the NaTRON format, eg the columns in the same order etc. 
This step includes fetching and attaching the correct *locationID* from the locations datatable.

```{r, eval = F}
myEvents <- str_map_events(data = setesdal, 
                conn = myConnection, 
                location_table = myLocationTable)

```




## 6. Upload structured event data to NaTron
This step uploads your structured event data table to NaTRON. For occurence datasets you may want to skipp this and the preceeding step.

```{r, eval = F}
upsert_events(myEvents, myConnection)

```



## 7. Structure occurrence-data
This step organises your occurrence data table to be compatible and ready to upload to NaTRON.

```{r, eval = FALSE}
myOccurences <- str_map_occ(setesdal, myConnection)

```
As of 10.09.2019 this produces a silent warning that *scientificName* has not been copied to the occurence table. Thta because it is not a column name in NaTRON. To fix this, or similar problems, modify the column name in the original data:
```{r, eval=FALSE}
colnames(setesdal)[colnames(setesdal) == "scientificName"] <- "organimsName"
myOccurences <- str_map_occ(setesdal, myConnection)


```


## 8. Upload structured occurence table to NaTRON
This step uploads your structured occurence table (step 7) to NaTRON.

```{r, eval=FALSE}
upsert_occ(myOccurences, myConnection)

```





## Examples of stylized data tables

#### *Example of wide data*

| Sampling date    | Sampling locations  | Alchemilla alpina   | Agrostis cappilaris | Andromeda polifolia |
| :-------------:  | :-------------:     | :-------------:     |:-------------:      | :-----:             |
| 2010             | A                   | 12                  | 6                   | 2                   |
| 2011             | A                   | 3                   | 17                  | 4                   |
| 2011             | B                   | 15                  | 2                   | 6                   |


#### *Example of long data*

| Sampling event    | Sampling date        | Sampling location   | Species       |      |  Quantity           |
| :-------------:   | :-------------:      |:-------------:      | :-------------:      |:-------------:      |
| 1                 | 2010                 |A                    |Alchemilla alpina    | 12                  |
| 1                 | 2010                 |A                    |Agrostis cappilaris  | 6                   |
| 1                 | 2010                 |A                    |Andromeda polifolia  | 2                   |
| 2                 | 2011                 |A                    |Alchemilla alpina    | 3                   |
| 2                 | 2011                 |A                    |Agrostis cappilaris  | 17                  |
| 2                 | 2011                 |A                    |Andromeda polifolia  | 4                   |
| 3                 | 2011                 |B                    |Alchemilla alpina    | 15                  |
| 3                 | 2011                 |B                    |Agrostis cappilaris  | 2                   |
| 3                 | 2011                 |B                    |Andromeda polifolia  | 6                   |

#### *Example of location table*

| locationID        | Sampling location | Coordinates 
|:-------------:    |:-------------:    |:-------------:
|a                  |A                  | x.xx 
|a                  |A                  | x.xx
|a                  |A                  | x.xx
|a                  |A                  | x.xx
|a                  |A                  | x.xx
|a                  |A                  | x.xx
|b                  |B                  | x.xx
|b                  |B                  | x.xx
|b                  |B                  | x.xx


#### *Example of event table*

| eventID    | Sampling event    | Sampling date        | locationID   |     
| :-------------:   | :-------------:   | :-------------:      |:-------------:     
| 1*                 | 1                 | 2010                 |a                   
| 1*                 | 1                 | 2010                 |a                   
| 1*                 | 1                 | 2010                 |a                   
| 2*                 | 2                 | 2011                 |a                   
| 2*                 | 2                 | 2011                 |a                   
| 2*                 | 2                 | 2011                 |a                   
| 3*                 | 3                 | 2011                 |b                   
| 3*                 | 3                 | 2011                 |b                   
| 3*                 | 3                 | 2011                 |b                   


#### *Example of occurence table*

| eventID           | Species       |      |  Quantity           |
| :-------------:   | :-------------:      |:-------------:      |
| 1*                |Alchemilla alpina    | 12                  |
| 1*                |Agrostis cappilaris  | 6                   |
| 1*                |Andromeda polifolia  | 2                   |
| 2*                |Alchemilla alpina    | 3                   |
| 2*                |Agrostis cappilaris  | 17                  |
| 2*                |Andromeda polifolia  | 4                   |
| 3*                |Alchemilla alpina    | 15                  |
| 3*                |Agrostis cappilaris  | 2                   |
| 3*                |Andromeda polifolia  | 6                   |


